% Frequently Asked Questions
% Copyright (C) Florian Negele

% This file is part of the Eigen Compiler Suite.

% Permission is granted to copy, distribute and/or modify this document
% under the terms of the GNU Free Documentation License, Version 1.3
% or any later version published by the Free Software Foundation.

% You should have received a copy of the GNU Free Documentation License
% along with the ECS.  If not, see <https://www.gnu.org/licenses/>.

\input{utilities}

\startchapter{Questions and Answers}{Frequently Asked Questions}{faq}
{This \documentation{} answers frequently asked questions about the \ecs{} and how to use its tools to accomplish common tasks.}

\newcommand{\question}[1]{\subsection*{#1}}

\section{General Questions}

This section assembles questions frequently asked about the \ecs{} in general.

\question{What is the \ecs{}? Why does it exist?}

The \ecs{} is a completely self-contained collection of software development tools.
It exists to be recognized and adopted as a free development toolchain which is hopefully as useful and easy to use as its source code is intended to be approachable and comprehensible for developers and students wanting to learn, maintain, and customize a complete toolchain.

\question{What does the name \ecs{} mean?}

The \ecs{} provides all essential freedoms guaranteed by free software and is maintainable by a single person.
The term eigen in its name thus means that the \ecs{} is literally your own, regardless of whether you are a user or a developer.

\question{Does the \ecs{} support Unicode source files?}

The \ecs{} currently only supports the ASCII character encoding.
It will however eventually support Unicode source files encoded in UTF-8.

\section{How-Tos}

This section describes how to accomplish commonly requested tasks using the tools of the \ecs{} and therefore assumes some basic knowledge of its features and functionality.

\question{How to disassemble plain binary files?}

Since disassemblers can only process object files which are usually generated by other tools of the \ecs{}, a plain binary file must first be converted into an object file.
The simplest way to achieve this conversion is to assemble an assembly source file consisting of a single line of code which just copies the contents of the binary file using the embed binary file directive.

\concludechapter
