% Common user interface
% Copyright (C) Florian Negele

% This file is part of the Eigen Compiler Suite.

% Permission is granted to copy, distribute and/or modify this document
% under the terms of the GNU Free Documentation License, Version 1.3
% or any later version published by the Free Software Foundation.

% You should have received a copy of the GNU Free Documentation License
% along with the ECS.  If not, see <https://www.gnu.org/licenses/>.

\input{utilities}

\startchapter{User Interface}{User Interface}{interface}
{The \ecs{} features a variety of different tools like compilers, assemblers, and linkers.
This \documentation{} describes their input and output and explains the common user interface that is shared across all of these tools.}

\epigraph{The trivial round, the common task, \\ would furnish all we ought to ask.}{John Keble}

\section{Introduction}

The \ecs{} is a toolchain that consists of several different programming tools targeting a variety of programming languages, hardware architectures, and runtime environments.
The following list categorizes all tools provided by the \ecs{} and visualizes their input and output:

\begin{itemize}

\item Preprocessors\index{Preprocessors}\nopagebreak

Preprocessors are tools that perform simple lexical substitutions like macro expansions and conditional inclusion on source code written in a programming language.
This very first phase of translating the source code is also part of all subsequent tools if the corresponding programming language supports preprocessing.
An example of a preprocessor tool is \tool{cpp\-prep}.

\flowgraph{\resource{source code} \ar[r] & \converter{Preprocessor} \ar[r] & \resource{preprocessed\\source code}}

\item Pretty Printers\index{Pretty printers}\nopagebreak

Pretty printers are tools that reformat source code written in a programming language by printing it again using a consistent layout.
Examples of pretty printer tools are \tool{asm\-print} and \tool{fal\-print}.

\flowgraph{\resource{source code} \ar[r] & \converter{Pretty Printer} \ar[r] & \resource{reformatted\\source code}}

\item Semantic Checkers\index{Semantic checkers}\nopagebreak

Semantic checkers perform syntactic and semantic checks on source code written in a programming language and print diagnostic messages.
Some checkers may have to process additional semantic information stored in separate files.
This phase of translating the source code is also part of all subsequent tools.
Examples of semantic checker tools are \tool{cpp\-check} and \tool{cd\-check}.

\flowgraph{\resource{source code} \ar[r] & \converter{Semantic\\Checker} \ar@/l/[d] \ar[r] & \resource{diagnostic\\messages} \\ & \resource{external\\information} \ar@/r/[u]}

\item Serializers\index{Serializers}\nopagebreak

Serializers are debugging tools that dump the complete internal representation of a program in a human readable format.
This potentially also allows external development tools to make use of the same internal program representation.
Examples of serialization tools are \tool{ob\-dump} and \tool{fal\-dump}.

\flowgraph{\resource{source code} \ar[r] & \converter{Serializer} \ar[r] & \resource{internal\\representation}}

\item Interpreters\index{Interpreters}\nopagebreak

Interpreters read and process source code for a programming language by emulating a runtime environment for this language and executing the program accordingly.
The actual interpreted program defines the input and output the interpreter processes and produces.
Examples of interpreter tools are \tool{cpp\-run} and \tool{cd\-run}.

\flowgraph{\resource{source code} \ar[r] & \converter{Interpreter} \ar@/u/[r] & \resource{input/\\output} \ar@/d/[l]}

\item Transpilers\index{Transpilers}\nopagebreak

Transpilers translate source code written in one programming language into source code for another programming language.
They store the resulting code in corresponding source code files.
Examples of transpiler tools are \tool{fal\-cpp} and \tool{ob\-cpp}.

\flowgraph{\resource{source code} \ar[r] & \converter{Transpiler} \ar[r] & \resource{translated\\source code}}

\item Documentation Generators\index{Documentation generators}\nopagebreak

Documentation generators extract the structure and comments of the source code in order to generate documentations about the program.
They store the resulting documentation in document files of various formats. \seedocumentation
Examples of documentation generator tools are \tool{cpp\-html} and \tool{ob\-latex}.

\flowgraph{\resource{source code} \ar[r] & \converter{Documentation\\Generator} \ar[r] & \resource{formatted\\document}}

\item Compilers\index{Compilers}\nopagebreak

Compilers translate programs written in a programming language into machine code for some hardware processors.
They store the resulting binary representation of the source code in object files. \seeobject
In addition, they also create debugging information and assembly code listings of the generated machine code.
Examples of compiler tools are \tool{cpp\-amd64} and \tool{ob\-avr}.

\flowgraph{\resource{source code} \ar[r] & \converter{Compiler} \ar[r] \ar[d] \ar[rd] & \resource{object file} \\ & \resource{debugging\\information} & \resource{assembly\\listing}}

\item Debugging Information Converters\index{Debugging information converters}\nopagebreak

Debugging information converters process debugging information files and generate debugging object files storing a binary representation thereof in a format suitable for the debugger of a specific runtime environment. \seedebugging
An example of a debugging information converter tool is \tool{dbg\-dwarf}.

\flowgraph{\resource{debugging\\information} \ar[r] & \toolbox{Converter} \ar[r] & \resource{debugging\\object file}}

\item Assemblers\index{Assemblers}\nopagebreak

Assemblers translate assembly code for some specific hardware architecture into equivalent binary machine code.
They store the resulting binary representation of the source assembly code in object files. \seeassembly
Examples of assembler tools are \tool{arma32\-asm} and \tool{ppc\-asm}.

\flowgraph{\resource{assembly\\source code} \ar[r] & \converter{Assembler} \ar[r] & \resource{object file}}

\item Disassemblers\index{Disassemblers}\nopagebreak

Disassemblers translate machine code for some specific hardware architecture stored in object files by printing it using a human-readable disassembly listing.
Examples of disassembler tools are \tool{mibl\-dism} and \tool{arma64\-dism}.

\flowgraph{\resource{object file} \ar[r] & \converter{Disassembler} \ar[r] & \resource{disassembly\\listing}}

\item Linkers\index{Linkers}\nopagebreak

Linkers assemble the binary representation of a program stored in object files to generate output files that are executable on some target platforms.
Examples of linker tools are \tool{link\-bin} and \tool{link\-prg}.

\flowgraph{\resource{object files} \ar[r] & \converter{Linker} \ar[r] & \resource{executable\\binary image}}

\end{itemize}

Although all of these tools process and produce different kinds of input and output, all of them present the same command-line interface to the user which is described in the following sections.

\section{Command-Line Arguments}\index{Command-line arguments}

Each tool of the \ecs{} provides a command-line user interface and accepts zero, one or more command-line arguments.
If there are one or more command-line arguments, each of them is treated as the name of an input file relative to the current working directory.
The notion of command-line options is not supported in order to make the contents of output files independent from the tool invocation.

Input files are plain text files containing the source code or other contents to be processed.
Output files are generated in the current working directory and have the same name as the input file being processed whereas the filename extension gets replaced by an appropriate suffix.
If the user does not provide any command-line argument, the tool reads the source code from the standard input stream.
Beforehand, it prints its version and a short copyright notice.

If there are more than one command-line argument, each input file is processed in sequence according to the given order of the arguments.
If there are errors during the processing of an input file, subsequent command-line arguments are ignored.
Some tools like assemblers and disassemblers do not depend on other input except for the actual source code.
In this case, the order of the command-line arguments is irrelevant and it would be equivalent to execute the tool once for every input file.

However, there are tools that depend on the actual order of the command-line arguments.
They even may behave differently if executed several times with the same command-line argument.
This may especially be the case for compilers which may store additional semantic information in separate files for every input file they compile.
In addition, interpreters may also depend on the actual order of the input files they have processed so far.

On the other hand, it is crucial for some tools to process several input files during the same execution of the tool.
Particularly linkers have to collect the information stored in several object files to generate an executable binary image.
In this case, the order of the arguments is not important, since all information is collected first before the linking process is started.
However, linkers generate output files named after the very first argument with the filename extension being replaced by an appropriate suffix.

\section{Diagnostic Messages}\index{Diagnostic messages}\index{Messages|see{Diagnostic messages}}

All tools of the \ecs{} use the same scheme for the diagnostic messages they generate.
The following list describes the different types of messages used by the \ecs{}:

\begin{itemize}

\item Errors\index{Error messages}\nopagebreak

Error messages indicate a problem within the contents of an input file that prohibits the tool from proceeding successfully.
Compilers for example use error messages to output syntactic or semantic errors that have to be fixed by the programmer.
In general, tools diagnose only the first few of the encountered problems but always yield a return code indicating the failure.

\item Fatal Errors\index{Fatal error messages}\nopagebreak

Fatal errors indicate a general failure of the tool that cannot be fixed by changing the contents of the input file.
This includes for example internal program errors, failures to open files, or critical system conditions like out of memory.

\item Warnings\index{Warnings}\nopagebreak

Warning messages indicate potential flaws within the contents of an input file.
They do not cause the tool to fail but they identify issues that may lead to unexpected behavior.
Therefore, warnings may be prevented by changing the contents of an input file without changing its semantics.

\item Notes\index{Notes}\nopagebreak

Notes give additional information about the result of processing an input file.
They are often used for debugging purposes or to diagnose violations of coding conventions.

\end{itemize}

All diagnostic messages are written to the standard error stream.
Each message consists of a short line of text and the name of the input file or the tool that caused the diagnostic message.
The message also contains the position\index{Position}\index{Diagnostic messages!Position} within the input file, if the problem can be located.
Since input files are text files in general, the position is usually given as the number of the text line containing the problem followed by the column therein.
For convenience, some tools like compilers also indicate the problematic position by reproducing the respective line of text from the input file.

\section{The \ecs{} Driver}\index{Eigen Compiler Suite!Driver}\index{Driver, Eigen Compiler Suite}

The \ecs{} provides a utility tool called \tool{ecsd} which autonomously drives its toolchain in order to build complete executable files for a specific runtime environment.
It infers the set of required tools like compilers, assemblers, and linkers from the type of its input files and automatically invokes one tool after the other with appropriate command-line arguments and environment variables\index{Environment variables}.

\flowgraph{\resource{input\\files} \ar[d] \ar[rd]|\hole \ar[rrd]|!{[r];[d]}\hole|\hole \ar@{-->}[r] & \converter{ecsd\vphantom{Compilers}} \ar@{~}[ld] \ar@{~}[d] \ar@{~}[rd] \ar@{-->}[r] & \resource{executable\\file} \\
\converter{Compilers} \ar[d] \ar[rd] & \converter{Assemblers\vphantom{Compilers}} \ar[d] & \converter{Linkers\vphantom{Compilers}} \ar[u] \\
\resource{debugging\\information\vphantom{support}} & \resource{object\vphantom{debugging}\\files\vphantom{support}} \ar[ru] & \resource{runtime\vphantom{debugging}\\support} \ar[u]}

In contrast to all other tools of the \ecs{}, this utility tool does support command-line options which influence how tools are identified and invoked.
The following list describes the form and behavior of the most important options:

\begin{description}

\item\syntax{"-b" <directory> $\mid$ "-""-base" <directory>}\alignright Use Specified Base Directory\nopagebreak

Allows overriding the base directory provided by the \environmentvariable{ECSBASE} environment variable which is used to locate tools and the necessary runtime support.
If omitted, the tool infers the base directory from its own location.

\item\syntax{"-c" $\mid$ "-""-compile"}\alignright Compile and Assemble Only\nopagebreak

Compiles and assembles input files without invoking the linker at the end.
The output is a set of object files rather than an executable file.

\item\syntax{"-d" $\mid$ "-""-disassemble"}\alignright Disassemble Object Files\nopagebreak

Invokes the disassembler instead of the linker at the end.
The output is a set of disassembly listings rather than an executable file.

\item\syntax{"-g" $\mid$ "-""-generate"}\alignright Generate Debugging Information\nopagebreak

Generates and includes debugging information during linking.
The resulting executable file can be executed and tested using a symbolic debugger.

\item\syntax{"-h" $\mid$ "-""-help"}\alignright Print Command-Line Help\nopagebreak

Prints information about all supported command-line arguments.
This especially also includes all supported source types and available target environments.

\item\syntax{"-i" <tool> $\mid$ "-""-invoke" <tool>}\alignright Invoke Specified Tool Only\nopagebreak

Processes either input files or the standard input stream using the specified tool.
The output is whatever that tool generates rather than an executable file.
\ifbook See page~\pageref{idx:tools} for an index of all available tools. \fi

\item\syntax{"-l" $\mid$ "-""-library"}\alignright Create Library File\nopagebreak

Combines input files into a single object file.
The output is a library file rather than an executable file.

\item\syntax{"-o" <format> $\mid$ "-""-output" <format>}\alignright Output Specified File Format\nopagebreak

Specifies the output file format of the linker.
The available file formats correspond to the suffixes of the linker tools listed in \Documentation{}~\documentationref{object}{Object File Representation}.
If omitted, the tool uses the default file format of the target environment.

\item\syntax{"-p" $\mid$ "-""-protect"}\alignright Protect Environment Variables\nopagebreak

Prevents the environment from being automatically modified when invoking tools that depend on environment variables.

\item\syntax{"-r" <support> $\mid$ "-""-runtime" <support>}\alignright Include Specified Runtime Support\nopagebreak

Includes additional runtime support stored in the specified object or library file during linking.
\ifbook See page~\pageref{idx:runtime} for an index of all available runtime support. \fi

\item\syntax{"-s" <type> $\mid$ "-""-source" <type>}\alignright Use Specified Source Type\nopagebreak

Allows defining the source type of input files for which the corresponding tool could not be inferred from the filename extension.

\item\syntax{"-t" <environment> $\mid$ "-""-target" <environment>}\alignright Target Specified Environment\nopagebreak

Specifies the target environment for cross compilations.
\ifbook See page~\pageref{idx:environment} for an index of all available target environments. \fi
If omitted, the tool tries to target its own runtime environment.

\item\syntax{"-v" $\mid$ "-""-verbose"}\alignright Print Command Line\nopagebreak

Enables verbose mode which prints the actual command line before invoking any tool.
This is helpful for examining how and which tools are invoked and what runtime support is provided.

\end{description}

The \tool{ecsd} driver tool is most useful when it is accessible via the \environmentvariable{PATH} environment variable of the runtime environment and is part of an appropriate installation of the \ecs{}.
See \Documentation{}~\documentationref{guide}{User Guide} for some examples of using this utility tool in practice.

\concludechapter
